---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# moon

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/DaCHE-SDU/moon/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/DaCHE-SDU/moon/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**moon** is an R implementation of the MOON (Modeling Obesity in Norway)
Markov cohort model originally published by Bjørnelv et al. (2021). It
simulates obesity prevalence, incremental health-care costs, and years
of life lost in a Norwegian birth cohort over the life course, and
supports both deterministic runs and 1000-iteration probabilistic
sensitivity analysis (PSA) with the published correlated-draw
conventions for mortality hazard ratios, costs, and transition
probabilities.

> **Status: early development.** `moon` is in active early development.
> The public API may change without deprecation warnings until the first
> tagged release, and outputs should be verified against your own
> expectations (see the [Disclaimer](#disclaimer)).

## Installation

`moon` is not on CRAN. Install the development version from GitHub:

``` r
# install.packages("pak")
pak::pak("DaCHE-SDU/moon")
```

## Quick example

```{r quick-example}
library(moon)

# 1. Build a Norwegian default `params` list (single sex per call).
params <- moon_params_norway(sex = "female")

# 2. Run the deterministic simulation.
run <- moon_deterministic(params)

run

# 3. Inspect costs and prevalence.
moon_costs(run, by = "state")
moon_prevalence(run, ages = 45)
```

For probabilistic uncertainty, use the spec'd parameter list and
`moon_psa()`:

```{r psa-example, eval = FALSE}
spec <- moon_params_norway(sex = "female", uncertainty = TRUE)
psa <- moon_psa(spec, n_iter = 1000, seed = 1)

summary(psa)
plot(psa) # forest plot of headline metrics
```

The PSA call materialises 1000 parameter draws via `moon_sample_params()`
and runs `moon_deterministic()` over each. See `?moon_psa` for the
correlation flags and `store_traces` options.

## Citation

If you use `moon` in published work, please cite the original paper that
specifies the model:

> Bjørnelv GMW, Halsteinli V, Kulseng BE, Sonntag D, Ødegaard RA.
> Modeling Obesity in Norway (The MOON Study): A Decision-Analytic
> Approach—Prevalence, Costs, and Years of Life Lost. *Medical Decision
> Making.* 2021;41(1):21–36.
> [doi:10.1177/0272989X20971589](https://doi.org/10.1177/0272989X20971589)

## Differences from the published paper

Running `moon` with the bundled Norwegian parameters produces results
that differ slightly from those reported in Bjørnelv et al. (2021),
due to minor improvements made to the model during its transition from
the original Excel/VBA implementation to R. The improvements are:

- **Mortality–cycle alignment.** Baseline life-table mortality is
  aligned so that each cycle uses the mortality probability for the age
  the cohort occupies at the start of that cycle (cycle 1, advancing
  from age 2 to 3, uses `q_2`).
- **Survival-conditional transitions.** State transitions are applied
  only to individuals who survive the cycle, so mortality and
  BMI-state transitions are modelled as sequential rather than
  independent events within a cycle.
- **Consistent survival functions across sex strata.** The same
  parametric survival distribution is used for each transition and age
  band across the female, male, and combined-sex models.

## Disclaimer

`moon` is a population-level decision-analytic model intended for
research and educational use. It is designed to inform exploratory
cohort-level analyses of obesity prevalence, health-care costs, and
years of life lost, and is **not** intended to model individual
patient trajectories. Results from this package should not be used as
the sole basis for health-policy, reimbursement, or regulatory
decisions without independent expert review and validation.

Users are responsible for verifying that the model's assumptions,
parameters, and outputs are appropriate for their intended use, and
for checking the correctness of any results produced by the package
in their analytical context. The authors and contributors accept no
responsibility for decisions made on the basis of model outputs.

The software is distributed under the MIT License, which disclaims
all warranties — see [LICENSE.md](LICENSE.md) for the full text.

## Contributing

Contributions are welcome. Please see the
[contributing guide](.github/CONTRIBUTING.md) for how to file issues and
open pull requests. Note that the `moon` project is released with a
[Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md); by contributing
you agree to abide by its terms.

## License

MIT. See [LICENSE](LICENSE) for the short form and
[LICENSE.md](LICENSE.md) for the full text.
