---
title: "Getting started with MOON"
subtitle: "Run the Norwegian baseline end-to-end: deterministic, scenarios, PSA"
author: "Nicolai Damslund"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    fig-width: 7
    fig-height: 4.5
    theme: cosmo
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Getting started with MOON}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(moon)
library(ggplot2)
```

# What this vignette covers

This is the entry-point tour of the **moon** package: how to run the
published Norwegian model with the bundled inputs and inspect what comes
back. Everything here uses defaults — no parameter editing.

If you want to *modify* the inputs (different costs, tighter mortality
HRs, swap the life table, build your own PSA specs), see the companion
vignette **"Customizing MOON"**.


# 1. The MOON model in one screen

MOON is a **discrete-time Markov cohort model** that simulates a
Norwegian birth cohort from age 2 to ~age 100, in 1-year cycles. The
state space is five live BMI bands plus an absorbing dead state:

- `N_always` — always normal weight (BMI 18.5–25)
- `N_prev` — currently normal weight but previously OW or OB
- `OW` — overweight (BMI 25–30)
- `OB1` — obese grade 1 (BMI 30–35)
- `OB2` — obese grade 2+ (BMI ≥ 35)
- `dead` — absorbing

Transition probabilities come from parametric survival analyses on
Norwegian longitudinal cohorts; mortality is age- and BMI-specific
(life table × hazard ratios from the Global BMI Mortality Collaboration);
costs are per-capita 2009 EUR/year by age and state. The original model
was implemented in Excel — see Bjørnelv et al. 2021,
*Medical Decision Making*,
[doi:10.1177/0272989X20971589](https://doi.org/10.1177/0272989X20971589).
This R port lifts every input into typed objects, adds an S3 layer over
the run output, and exposes a clean PSA path through a parameter-spec
system.


# 2. The public surface

The public API is deliberately compact:

| Function | Purpose |
|---|---|
| `moon_params_norway(sex, uncertainty)` | Build the Norwegian-baseline parameter list (deterministic or PSA-ready). |
| `moon_check_params(params)` | Validate shapes / row sums / value ranges. |
| `moon_deterministic(params, tp_overrides)` | One Markov run; returns a `moon_deterministic`. |
| `moon_psa(spec, n_iter, seed, ...)` | A full PSA; returns a `moon_psa`. |
| `moon_prevalence()`, `moon_costs()` | Long-form extractors on a run object. |
| `print()` / `summary()` / `plot()` / `as.data.frame()` | S3 methods on `moon_deterministic` and `moon_psa`. |

The minimum end-to-end example is three lines:

```{r minimal-example}
params <- moon_params_norway("female")
res <- moon_deterministic(params)
summary(res)
```

Everything else in this document is layering on top of these calls.


# 3. Loading parameters

`moon_params_norway()` builds the engine-shape `params` list for the
published Norwegian birth cohort. Pass one sex per call:

```{r load-params}
params_F <- moon_params_norway("female")
str(params_F, max.level = 1)
```

Slot anatomy:

| Slot | Type | What it carries |
|---|---|---|
| `start_age`, `max_age` | int | model horizon (2 → 100) |
| `discount_rate` | num | 0.04 by default |
| `cost_currency` | chr | "EUR" |
| `cohort_n` | named int | cohort size, key = sex label |
| `init_prev` | named num | NW/OW/OB1/OB2 mass at `start_age` |
| `qx` | named num | NW baseline yearly mortality, ages 2–100 |
| `mortality_hr` | data.frame | HRs vs NW for OW/OB1/OB2 by age band |
| `transition_probs` | data.frame | yearly between-state transitions, ages 2–99 |
| `cost_df` | data.frame | per-capita yearly cost by age × state |

`cohort_n` is a *named* integer (`c(female = 26458L)`) so the sex label
is recoverable from the params alone — useful when you stitch results
from multiple sexes together.

Setting `uncertainty = TRUE` returns the same shape but with
the uncertain slots wrapped as `moon_param_*` spec objects, ready for
PSA. We come back to that in [§11](#psa-on-the-norwegian-baseline).


# 4. Validating with `moon_check_params()`

```{r validate}
moon_check_params(params_F)
```

`moon_check_params()` is silent on success and errors loudly on
problems: shape mismatches, off-simplex `init_prev`, transition
probabilities outside [0, 1], age misalignment between
`transition_probs` and `start_age`/`max_age`, etc. Three phases run in
order — structural, range, then cross-object consistency — and any
problems found are reported in one batch.

It runs implicitly inside `moon_deterministic()` (with `strict = TRUE`).
Calling it standalone is useful when you've mutated `params` and want to
fail fast before kicking off a full run; pass `strict = FALSE` to
downgrade the error to a warning while iterating.


# 5. Running deterministically

```{r run-deterministic}
res_F <- moon_deterministic(params_F)
class(res_F)
names(res_F)
```

A `moon_deterministic` is a list with four named slots:

- **`$trace`** — long data frame `(age, sex, state, n)`. Six states
  including `dead` and the engine's full `N_always` / `N_prev` split. `n`
  is a head-count denominated by `params$cohort_n`.
- **`$costs`** — long data frame `(age, sex, state, cost, cost_disc)`,
  five live states. `cost` is total euros that age-state cell incurred;
  `cost_disc` applies `params$discount_rate`. `dead` is excluded.
- **`$params`** — the input list, stored verbatim. Useful for auditing
  and re-running.
- **`$meta`** — runtime metadata (R version, timestamp, duration,
  horizon, discount rate, applied `tp_overrides`).

```{r preview-trace-costs}
head(res_F$trace, 6)
head(res_F$costs, 6)
```


# 6. Inspecting with S3 methods

## 6.1 `print()` — the headline

```{r print}
print(res_F)
```

## 6.2 `summary()` — richer metrics

`summary()` returns a `summary.moon_deterministic` (a list with a print
method). Capture it and pull fields out for downstream calculation:

```{r summary}
summary(res_F)

s <- summary(res_F)
s$LE
s$prev_age45
s$inc_cost_state
```

## 6.3 `plot()` — five canned views

`plot.moon_deterministic()` takes a `type =` argument with five
options. All return ggplot objects, so `+ theme_*()` and `ggsave()`
work without reaching inside.

::: {.panel-tabset}

### occupancy

```{r plot-occupancy}
plot(res_F, type = "occupancy")
```

Cohort head-counts by state. The `dead` line rises monotonically.

### prevalence_alive

```{r plot-prevalence-alive}
plot(res_F, type = "prevalence_alive")
```

Each state as a fraction of the cohort still alive at that age. Lines
sum to 1 per age.

### prevalence_initial

```{r plot-prevalence-initial}
plot(res_F, type = "prevalence_initial")
```

Each state as a fraction of the **initial** cohort. Lines no longer
sum to 1 because `dead` siphons mass over time — useful when you want
survivor-adjusted prevalences directly.

### survival

```{r plot-survival}
plot(res_F, type = "survival")
```

Kaplan–Meier-style survival curve.

### costs

```{r plot-costs}
plot(res_F, type = "costs")
```

Per-capita per-age incremental cost vs an "everyone is NW"
counterfactual, broken out by state.

:::

## 6.4 `as.data.frame()`

```{r as-data-frame}
head(as.data.frame(res_F, what = "trace"), 3)
head(as.data.frame(res_F, what = "costs"), 3)
```

Identical to `res_F$trace` / `res_F$costs` but reads more idiomatically
in pipelines.


# 7. Long-form extractors

The S3 plots cover the headlines; most real analyses want raw long-form
data. There are two public extractors plus the underlying `$trace` /
`$costs` frames.

## 7.1 `moon_prevalence()`

```{r prev-alive}
moon_prevalence(res_F, denominator = "alive", ages = c(2, 30, 45, 65))
```

Same data with the **initial** denominator (note the `dead` row at age
65, and that the within-age sum is no longer 1):

```{r prev-initial}
moon_prevalence(res_F, denominator = "initial", ages = c(2, 30, 45, 65))
```

## 7.2 `moon_costs()`

`moon_costs()` aggregates `$costs` over a single dimension chosen via
`by =` (`"total"` returns a scalar; `"age"`, `"state"`, or `"sex"`
return a long data frame). Restrict ages with `ages =`, switch to
discounted euros with `discounted = TRUE`.

```{r costs-total}
moon_costs(res_F, by = "total")
moon_costs(res_F, by = "total", discounted = TRUE)
```

```{r costs-by-state}
moon_costs(res_F, by = "state")
```

```{r costs-by-age}
moon_costs(res_F, by = "age", discounted = TRUE, ages = c(20, 50, 80))
```


# 8. Building your own plot

A worked example: take the full cohort decomposition out of
`moon_prevalence()` and turn it into a stacked-area plot — every state
on one axis, summing to 100% of the initial cohort at every age.

```{r custom-plot}
state_mix <- moon_prevalence(res_F, denominator = "initial")
state_mix$state <- factor(
  state_mix$state,
  levels = c("dead", "N_always", "N_prev", "OW", "OB1", "OB2")
)

state_palette <- c(
  dead     = "grey70",
  N_always = "#1b7837",
  N_prev   = "#7fbf7b",
  OW       = "#f1a340",
  OB1      = "#e08214",
  OB2      = "#b35806"
)

ggplot(state_mix, aes(x = age, y = prevalence, fill = state)) +
  geom_area(position = "stack", colour = "white", linewidth = 0.1) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = c(0, 0)
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = state_palette) +
  labs(
    x = "Age",
    y = "Share of initial cohort",
    fill = NULL,
    title = "Female cohort state composition — Norwegian baseline",
    subtitle = "Stack reaches 100% of the age-2 cohort at every age"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

You are never locked into the canned plots. The extractors and
`$trace` / `$costs` frames give you raw long data; everything else is
whatever your downstream code does best.


# 9. Comparing sexes

`moon_params_norway()` is single-sex per call. Run all three strata and
stitch the headline numbers into one table:

```{r compare-sexes}
res_M <- moon_deterministic(moon_params_norway("male"))
res_both <- moon_deterministic(moon_params_norway("both"))

inc_cost_per_capita <- function(result) {
  m <- merge(
    result$trace,
    result$costs,
    by = c("age", "sex", "state"),
    all.x = TRUE
  )
  m$c <- ifelse(m$n > 0 & !is.na(m$cost), m$cost / m$n, 0)
  c_NW <- m$c[m$state == "N_always"]
  names(c_NW) <- m$age[m$state == "N_always"]
  m$c_NW <- c_NW[as.character(m$age)]
  obese <- m$state %in% c("OW", "OB1", "OB2")
  sum(m$n[obese] * (m$c[obese] - m$c_NW[obese])) /
    unname(result$params$cohort_n)
}

data.frame(
  sex = c("female", "male", "both"),
  inc_cost_eur = round(c(
    inc_cost_per_capita(res_F),
    inc_cost_per_capita(res_M),
    inc_cost_per_capita(res_both)
  ))
)
```

These are lifetime cumulative incremental costs per cohort member,
relative to an all-NW counterfactual.


# 10. Built-in scenarios via `tp_overrides=`

Every paper scenario can be run by passing a one-shot `tp_overrides`
list to `moon_deterministic()`. Three slots are recognised:

- `set_zero` — character vector of transition columns to zero out
  (e.g. `"OW_OB1"`, which closes the only path *into* the obesity
  ladder).
- `init_prev` — replace the params' initial state vector for this run.
- `start_age` — start the cohort later than `params$start_age` (used
  by paper scenarios SA3-6, which start everyone at age 30).

The example below reproduces paper scenario **SA2** (eliminate OB1+OB2
by closing OW→OB1 and re-allocating any starting OB1+OB2 mass to OW):

```{r sa2}
ip <- params_F$init_prev
ip_sa2 <- c(
  NW = unname(ip["NW"]),
  OW = unname(ip["OW"] + ip["OB1"] + ip["OB2"]),
  OB1 = 0,
  OB2 = 0
)

res_F_sa2 <- moon_deterministic(
  params_F,
  tp_overrides = list(set_zero = "OW_OB1", init_prev = ip_sa2)
)

c(
  LE_baseline = summary(res_F)$LE,
  LE_SA2 = summary(res_F_sa2)$LE,
  YLL_female = summary(res_F_sa2)$LE - summary(res_F)$LE
)
```

`YLL_female` is the per-person years-of-life lost attributable to
obesity in this cohort — the gap between the obesity-eliminated
counterfactual and the baseline run.

> `tp_overrides=` is a one-shot flag passed at the call site; the
> `params` object is untouched. If you instead want a permanent change
> ("this is my new baseline"), edit `params` directly — see the
> "Customizing MOON" vignette.


# 11. PSA on the Norwegian baseline

PSA uses the **same loader**, just with `uncertainty = TRUE`. The slots
the paper treats as random become `moon_param_*` spec objects;
everything else stays plain.

```{r psa-spec}
spec_F <- moon_params_norway("female", uncertainty = TRUE)
str(spec_F, max.level = 1)
```

The slot shapes match the deterministic loader, but the contents of
the uncertain slots have changed:

```{r psa-spec-classes}
class(spec_F$mortality_hr$OW[[1]])
class(spec_F$cost_df$cost[[1]])
class(spec_F$transition_probs$specs[[1]])
length(spec_F$transition_probs$specs)
```

- `mortality_hr$OW` (and `$OB1`, `$OB2`) is a list-column of
  `moon_param_lognormal` specs, one per age band.
- `cost_df$cost` is a list-column of `moon_param_gamma` specs, one
  per `(age, state)` cell. Ages 2–19 have mean = se = 0 (degenerate
  gammas that always sample 0).
- `transition_probs` becomes `list(specs, bands)` — `specs` holds 18
  `moon_param_mvnorm` objects (one per transition × age band).

`init_prev`, `qx`, scalars, and `cohort_n` stay as plain values: the
paper does not treat these as uncertain.

## 11.1 Running PSA

```{r psa-run}
psa_F <- moon_psa(
  spec_F,
  n_iter = 100,
  seed = 123,
  store_traces = "all"
)
class(psa_F)
names(psa_F)
```

`store_traces` controls how much you keep:

- `"none"` — discard per-iteration cohort traces. Smallest object;
  enough for forest plots.
- `"summary"` (default) — keep iteration-level summary metrics only.
- `"all"` — keep every iteration's full `moon_deterministic`. Required
  for the `incremental_cost_age` band plot. Uses ~few-hundred KB per
  iteration.

The `moon_psa` return has six slots:

- `$summary` — aggregated stats `(sex, metric, mean, lower95,
  upper95, sd)`. Twelve metrics per sex.
- `$per_iter` — long table `(iter, sex, metric, value)` — the source
  for any custom credible-interval calculation.
- `$traces` — depends on `store_traces`.
- `$draws` — long table of HR sampled values keyed by
  `(iter, parameter)`. Audit trail for "which exact draws produced
  this iteration's results".
- `$params_spec` — the input spec, so you can re-run from scratch.
- `$meta` — `n_iter`, `seed`, `parallel`, `runtime_sec`, the
  correlation flags, `store_traces`, `tp_overrides`.

```{r psa-frames}
head(psa_F$summary)
head(psa_F$per_iter)
```

## 11.2 Inspecting with S3 methods

```{r psa-print}
print(psa_F)
summary(psa_F)
```

`plot.moon_psa()` has two types:

::: {.panel-tabset}

### forest

```{r psa-plot-forest}
plot(psa_F, type = "forest")
```

Point estimate + 95% CI for every metric.

### incremental_cost_age

```{r psa-plot-cost-age}
plot(psa_F, type = "incremental_cost_age")
```

Per-age incremental cost vs NW with mean and 2.5–97.5% bands across
iterations. Requires `store_traces = "all"`.

:::

## 11.3 Custom credible intervals from `$per_iter`

Same "you're not locked in" message as [§8](#building-your-own-plot) — `$per_iter` is a long
data frame with one row per (iter × metric × sex), so any
quantile-based summary is one base-R call.

```{r psa-custom-ci}
le_draws <- psa_F$per_iter$value[psa_F$per_iter$metric == "LE"]
quantile(le_draws, c(0.025, 0.5, 0.975))

inc_total <- psa_F$per_iter$value[
  psa_F$per_iter$metric == "cum_inc_cost_total_undisc"
]
sprintf(
  "Female mean inc cost: EUR %s [EUR %s, EUR %s]",
  format(round(mean(inc_total)), big.mark = ","),
  format(round(quantile(inc_total, 0.025)), big.mark = ","),
  format(round(quantile(inc_total, 0.975)), big.mark = ",")
)
```

## 11.4 Reproducibility & speed knobs

- **`seed`** — sets the random seed before sampling. Same seed + same
  spec ⇒ bit-identical PSA results.
- **`correlate_hr` / `correlate_cost`** — toggle the legacy single-Z /
  single-U shared-randomness behaviour. Default `TRUE` matches what
  the paper authors ran. Set `FALSE` for independent draws per spec.
- **`parallel = TRUE`** — opt into `furrr` (if installed). Iterations
  parallelise trivially; skipped here so the demo renders deterministically.
- **`store_traces`** — set to `"summary"` or `"none"` for big runs;
  the saved object can otherwise grow to hundreds of MB at 1000 iterations.


# 12. PSA on a built-in scenario

Pass `tp_overrides=` to `moon_psa()` and the override is applied to
every iteration. This gives you a **credible interval on a scenario
quantity** — for SA2 (eliminate OB1+OB2), the YLL distribution.

```{r psa-scenario}
psa_F_base <- moon_psa(
  spec_F,
  n_iter = 100,
  seed = 123,
  store_traces = "none"
)
psa_F_sa2 <- moon_psa(
  spec_F,
  n_iter = 100,
  seed = 123,
  tp_overrides = list(set_zero = "OW_OB1", init_prev = ip_sa2),
  store_traces = "none"
)

le_base <- psa_F_base$per_iter$value[psa_F_base$per_iter$metric == "LE"]
le_sa2 <- psa_F_sa2$per_iter$value[psa_F_sa2$per_iter$metric == "LE"]
yll <- le_sa2 - le_base

sprintf(
  "Female YLL from eliminating OB1+OB2: %.2f y [%.2f, %.2f]",
  mean(yll),
  quantile(yll, 0.025),
  quantile(yll, 0.975)
)
```

Same `seed` for both runs ⇒ paired draws ⇒ the YLL distribution is
the within-iteration difference, not a difference of independent
samples — which is what you want for a credible interval on the
incremental quantity.


# 13. Where to go next

- **"Customizing MOON"** — companion vignette covering how to alter
  any input (deterministic and PSA), swap in your own datasets, and
  build PSA specs from scratch.
- **`?moon_deterministic`, `?moon_psa`, `?moon_params_norway`** —
  function-by-function reference.
- **`tests/testthat/`** — anchor values for the deterministic and
  PSA paths; the reference `.rds` files are the ground truth a
  build must reproduce to 1e-10.
- **Bjørnelv et al. 2021** —
  [doi:10.1177/0272989X20971589](https://doi.org/10.1177/0272989X20971589).
  The supplementary appendix has the per-band transition tables and
  the cost-model structure.
