---
title: "Customizing MOON"
subtitle: "Bring your own inputs: deterministic edits, full data swaps, custom PSA specs"
author: "Nicolai Damslund"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    fig-width: 7
    fig-height: 4.5
    theme: cosmo
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Customizing MOON}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(moon)
library(ggplot2)
```

# What this vignette covers

How to customize the MOON model: tweak any field of the `params` list,
swap in your own data files, or build PSA specs from scratch. It assumes
you've read the **"Getting started with MOON"** vignette and are
comfortable with the canonical `moon_params_norway()` →
`moon_check_params()` → `moon_deterministic()` workflow.


# 1. The mental model

`moon_params_norway()` returns a plain R list. Every slot is yours to
mutate before passing to `moon_deterministic()` (or
`moon_sample_params()` / `moon_psa()` on the spec'd path). There are
three escalating levels of intervention:

1. **`tp_overrides=`** at the call site. One-shot scenarios on top of an
   unchanged `params` (covered in the getting-started vignette [§10](moon-getting-started.html#built-in-scenarios-via-tp_overrides)).
2. **In-place mutation** of fields on `params`. New "baseline" for every
   downstream call. Most of this vignette.
3. **Whole-cohort swap** via the `data_dir =` argument or by
   constructing `params` from scratch. [§8](#bringing-your-own-data-via-data_dir)–[9](#building-params-from-scratch).

`moon_check_params()` enforces what it can — slot shapes, value ranges,
age alignment between `transition_probs` and `start_age`/`max_age`,
exact `mortality_hr` band structure (35 / 50 / 70), simplex `init_prev`,
and so on. It **cannot** check semantic correspondence: that your
`init_prev` vector matches the cohort age you've named `start_age`,
that your costs are in the currency you've labelled them, or that the
life table you've loaded is for the population you think you're
modelling. Those are your responsibility.

A useful pattern while iterating:

```{r check-non-strict, eval = FALSE}
moon_check_params(params, strict = FALSE)
```

`strict = FALSE` downgrades the error to a warning and returns `params`
unchanged, so you can see *all* the problems at once instead of fixing
them one error at a time.


# 2. Tweaking scalars

The simplest edits. `discount_rate`, `cost_currency`, and `cohort_n` are
all single-cell slots.

```{r scalars}
params <- moon_params_norway("female")

params$discount_rate <- 0
params$cohort_n <- c(female = 1L) # turn head-counts into proportions

res <- moon_deterministic(params)

c(LE = summary(res)$LE, total_disc = summary(res)$total_cost_disc)
```

Setting `cohort_n` to 1 makes `$trace$n` a per-capita probability mass
rather than a head-count. The model behaviour is unchanged — `cohort_n`
is purely a denominator.


# 3. Editing `transition_probs`

`transition_probs` is a data frame with one row per age in
`start_age:(max_age - 1L)` and one column per transition (`NW_OW`,
`OW_NW`, `OW_OB1`, `OB1_OW`, `OB1_OB2`, `OB2_OB1`).

## 3.1 Age-restricted multiplications

Halve every yearly NW→OW probability for childhood ages only:

```{r tp-edit}
params <- moon_params_norway("female")
idx <- params$transition_probs$age <= 17
params$transition_probs$NW_OW[idx] <-
  params$transition_probs$NW_OW[idx] * 0.5

res_low <- moon_deterministic(params)
```

Compare against the untouched baseline by stitching two long frames:

```{r tp-edit-plot}
res_base <- moon_deterministic(moon_params_norway("female"))

prev_base <- transform(
  moon_prevalence(res_base, "alive"),
  scenario = "Baseline"
)
prev_low <- transform(
  moon_prevalence(res_low, "alive"),
  scenario = "Halved NW->OW (ages 2-17)"
)
prev <- rbind(prev_base, prev_low)
prev <- prev[prev$state %in% c("N_always", "N_prev", "OW", "OB1", "OB2"), ]

ggplot(
  prev,
  aes(x = age, y = prevalence, colour = state, linetype = scenario)
) +
  geom_line(linewidth = 0.9) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    x = "Age",
    y = "Prevalence (% of alive)",
    title = "Effect of halving childhood NW->OW transitions",
    colour = "State",
    linetype = NULL
  ) +
  theme_minimal()
```

## 3.2 Replacing a whole column

The same applies to wholesale replacement — assign a new vector of the
same length:

```{r tp-replace, eval = FALSE}
params$transition_probs$OB1_OB2 <-
  pmin(params$transition_probs$OB1_OB2 * 1.25, 1)
```

The validator will catch an out-of-range value if you skip the `pmin()`
guard:

```
moon_check_params() found 1 problem(s):
  - `transition_probs$OB1_OB2` contains values outside [0, 1] or non-finite.
```

## 3.3 What you cannot change

The age column is fixed: `transition_probs$age` must be exactly
`start_age:(max_age - 1L)`. If you want to shrink the simulation
horizon, change `start_age` / `max_age` *and* slice
`transition_probs` to match — see [§6](#init_prev-and-the-start_age-coupling) for the coupled case with
`init_prev`.


# 4. Editing `cost_df`

`cost_df` is long: one row per `(age, state)` for ages 2..100 and the
four live states `NW` / `OW` / `OB1` / `OB2`. The validator enforces
that range regardless of `start_age` / `max_age`, so leave the row
inventory alone — only change values.

## 4.1 Inflating to a different price year

Adjust 2009 EUR to a later cost year with a single multiplier:

```{r cost-inflate}
params <- moon_params_norway("female")
params$cost_df$cost <- params$cost_df$cost * 1.30 # ~2009 -> 2024 (illustrative)

summary(moon_deterministic(params))$total_cost
```

## 4.2 Replacing one state's age curve

Set `OB2` per-capita yearly cost to 4,000 EUR for adults, 0 for
children, in one assignment:

```{r cost-replace}
params <- moon_params_norway("female")
ob2 <- params$cost_df$state == "OB2"
params$cost_df$cost[ob2] <- ifelse(
  params$cost_df$age[ob2] >= 20,
  4000,
  0
)
moon_check_params(params)
```


# 5. Editing `mortality_hr`

`mortality_hr` is a three-row data frame keyed by age band
(`age_lower = c(35, 50, 70)`) with columns `OW`, `OB1`, `OB2`. The
validator pins the band structure: you can change values, but not the
number or position of bands.

A useful sensitivity: decouple BMI from mortality entirely (set every
HR to 1):

```{r hr-decouple}
params <- moon_params_norway("female")
params$mortality_hr$OW <- 1
params$mortality_hr$OB1 <- 1
params$mortality_hr$OB2 <- 1

c(
  LE_baseline = summary(moon_deterministic(moon_params_norway("female")))$LE,
  LE_no_hr = summary(moon_deterministic(params))$LE
)
```

The gap is the cohort-level life-expectancy attributable to the
BMI-mortality coupling.


# 6. `init_prev` — and the `start_age` coupling

`init_prev` is the **BMI distribution at `start_age`**. The field name
doesn't say so, but the engine treats it that way: row 1 of the cohort
trace at age `start_age` is exactly `cohort_n * init_prev`. The
validator does not check that `init_prev` matches the age you've
named — that's on you.

This matters whenever you change `start_age`. The bundled `init_prev`
(`c(NW = 0.898, OW = 0.090, OB1 = 0.0086, OB2 = 0.0033)`) is the
published age-2 distribution. If you set `start_age = 4` and leave
`init_prev` alone, you're claiming the cohort still has the age-2
prevalence at age 4 — which it doesn't.

Two ways to do this honestly:

## 6.1 Bootstrap from a baseline run

Run the unmodified model first, read the prevalence at your target age
out of `moon_prevalence()`, and feed it back as `init_prev` for the
shorter run:

```{r init-prev-bootstrap}
res_base <- moon_deterministic(moon_params_norway("female"))

# Prevalence at age 4 among alive, collapsed to the four init_prev states
p4 <- moon_prevalence(res_base, denominator = "alive", ages = 4)
ip4 <- c(
  NW = p4$prevalence[p4$state == "N_always"] +
    p4$prevalence[p4$state == "N_prev"],
  OW = p4$prevalence[p4$state == "OW"],
  OB1 = p4$prevalence[p4$state == "OB1"],
  OB2 = p4$prevalence[p4$state == "OB2"]
)
ip4
```

Now build a `params` whose age window starts at 4, with this matched
`init_prev`. We also need to slice `transition_probs` to the new range:

```{r start-age-shift}
params4 <- moon_params_norway("female")
params4$start_age <- 4L
params4$init_prev <- ip4
params4$transition_probs <- params4$transition_probs[
  params4$transition_probs$age >= 4L,
]

moon_check_params(params4)
res4 <- moon_deterministic(params4)
summary(res4)$LE
```

Note: `qx` already tolerates extra ages (the validator only flags
*missing* ones), so the bundled life table doesn't need slicing.
`cost_df` is checked against the fixed 2..100 range, so it's fine
too. Only `transition_probs` requires an exact `start_age:(max_age - 1)`
match.

## 6.2 Bring your own age-N prevalence

If you have empirical age-4 BMI prevalence from a different source,
just assign it directly:

```{r init-prev-byo, eval = FALSE}
params4$init_prev <- c(NW = 0.85, OW = 0.12, OB1 = 0.02, OB2 = 0.01)
```

The validator only checks that it sums to 1 and has the right names —
the *correctness* of the values is your problem.


# 7. Editing `qx`

`qx` is the per-age NW-baseline yearly mortality probability. Names are
ages as character (`"2"`, `"3"`, ...), values are in [0, 1]. The
validator tolerates extra ages beyond `start_age:(max_age - 1L)`, so
swapping in a longer life table is fine; missing ages will fail.

```{r qx-swap}
params <- moon_params_norway("female")

# Sensitivity: 10% higher background mortality at every age
params$qx <- pmin(params$qx * 1.10, 1)

moon_check_params(params)
```

If you want to swap in a completely different country's life table, the
cleanest path is `data_dir =` (next section) — that lets the loader do
the column-name and age-filter work for you. Direct assignment is for
small in-place tweaks.


# 8. Bringing your own data via `data_dir=`

`moon_params_norway()` reads its CSVs from `data_dir`, which defaults
to the bundled `inst/extdata/` directory. Point it at a directory of
files with the same filenames and schemas, and you can swap the entire
parameter set without touching R.

Expected layout. Cost and transition filenames are sex-keyed and not
configurable; the life table file name is configurable via
`lifetable_file =` (defaulting to `"Lifetable_Norway_2017.csv"`):

| File | What it carries |
|---|---|
| `Lifetable_Norway_2017.csv` (or whatever you pass to `lifetable_file =`) | columns: `Age`, `F`, `M`, `Both` (semicolon-separated) |
| `Cost_Female_2_100.csv` (and `_Male_`, `_Both_`) | columns: `Age`, `State`, `Cost_Mean`, `Cost_SE` |
| `TransitionParamsFemale.csv` (and `Male`, `Both`) | columns: `age_start`, `age_end`, `survival_function_starting_age`, `dist`, `mean1`, `mean2`, `Transition`, plus covariance entries `cov_r1c1`, `cov_r1c2`, `cov_r2c1`, `cov_r2c2` (only used when `uncertainty = TRUE`) |

The CSV State value `"N"` is renamed to `"NW"` on read; transition
labels in the source files use `N_OW` / `OW_N` / etc. and are renamed
to engine names (`NW_OW` / `OW_NW`) at load time.

```{r data-dir, eval = FALSE}
my_dir <- "~/my-moon-inputs"
params_C <- moon_params_norway(
  sex = "female",
  data_dir = my_dir,
  lifetable_file = "Lifetable_Denmark_2020.csv"
)
moon_check_params(params_C)
```

The bundled `inst/extdata/` directory of the installed package is the
canonical schema; copy it as a starting point:

```{r extdata-path, eval = FALSE}
list.files(system.file("extdata", package = "moon"))
```


# 9. Building `params` from scratch

If you have nothing in CSV form — just R objects — you can construct
the list directly. The required slots and their types are listed in
the getting-started vignette [§3](moon-getting-started.html#loading-parameters); the validator will tell you what's
wrong as you iterate. Use `strict = FALSE` to see every problem at
once:

```{r build-from-scratch, eval = FALSE}
params <- list(
  start_age = 2L,
  max_age = 100L,
  discount_rate = 0.04,
  cost_currency = "EUR",
  cohort_n = c(female = 26458L),
  init_prev = c(NW = 0.898, OW = 0.090, OB1 = 0.009, OB2 = 0.003),
  qx = setNames(my_qx_vec, as.character(2:99)),
  mortality_hr = data.frame(
    age_lower = c(35, 50, 70),
    OW = ...,
    OB1 = ...,
    OB2 = ...
  ),
  transition_probs = data.frame(
    age = 2:99,
    NW_OW = ...,
    OW_NW = ...,
    OW_OB1 = ...,
    OB1_OW = ...,
    OB1_OB2 = ...,
    OB2_OB1 = ...
  ),
  cost_df = data.frame(
    age = rep(2:100, times = 4),
    state = rep(c("NW", "OW", "OB1", "OB2"), each = 99),
    cost = ...
  )
)

moon_check_params(params, strict = FALSE) # iterate until clean
```


# 10. PSA: replacing a single spec

`moon_params_norway("...", uncertainty = TRUE)` returns the same shape
as the deterministic loader but with the uncertain slots replaced by
`moon_param_*` spec objects. To customize one of them, build a new
spec and assign it back into the list-column.

A worked example: the OW mortality HR for ages 35–50 has a published
95% CI of [1.15, 1.20]. Suppose you want to widen this to a more
conservative [1.05, 1.30]:

```{r psa-replace-spec}
spec <- moon_params_norway("female", uncertainty = TRUE)

spec$mortality_hr$OW[[1]] <- moon_param_lognormal(
  point = 1.17,
  lower = 1.05,
  upper = 1.30
)

# Quick sanity check on the new spec
moon_param_value(spec$mortality_hr$OW[[1]])
set.seed(1)
moon_param_sample(spec$mortality_hr$OW[[1]], n = 5)
```

That's all that's needed — `moon_psa(spec, ...)` will pick up the
new spec the next time it samples.


# 11. PSA: building specs from scratch

The constructors:

| Constructor | When to use |
|---|---|
| `moon_param_fixed(value)` | Pin a parameter you don't have uncertainty for. |
| `moon_param_lognormal(point, lower, upper)` | Hazard ratios; positive multiplicative quantities with published 95% CIs. |
| `moon_param_gamma(mean_vec, se_vec)` | Costs (positive, right-skewed). |
| `moon_param_mvnorm(mean_vec, cov_mat, dist, cycles)` | Survival-model coefficient vectors with covariance. |
| `moon_param_dirichlet(alpha)` | Simplex-valued vectors (e.g. an uncertain `init_prev`). |

Every spec implements two generics:

- `moon_param_value(x)` — the deterministic point estimate. Used as
  a safety net when a spec is encountered in a deterministic context.
- `moon_param_sample(x, n, ...)` — `n` random draws.

```{r constructors}
fix <- moon_param_fixed(1.5)
ln <- moon_param_lognormal(point = 1.45, lower = 1.30, upper = 1.62)
gam <- moon_param_gamma(mean_vec = c(0, 100, 250), se_vec = c(0, 30, 60))
dir <- moon_param_dirichlet(c(NW = 90, OW = 9, OB1 = 0.7, OB2 = 0.3))

set.seed(42)
moon_param_value(ln)
moon_param_sample(ln, n = 5)

moon_param_value(dir)
moon_param_sample(dir, n = 3)
```

Drop a fully-custom Dirichlet `init_prev` into a spec and you've moved
from "fixed entry distribution" to "uncertain entry distribution":

```{r dirichlet-init-prev, eval = FALSE}
spec$init_prev <- moon_param_dirichlet(
  c(NW = 89.8, OW = 8.98, OB1 = 0.86, OB2 = 0.33)
)
```

(Note: `moon_psa()` does not currently treat `init_prev` as uncertain
because the published model doesn't — see the `moon_sample_params()`
source if you want to extend the sampler.)


# 12. Shared-randomness flags

`moon_psa()` has two correlation toggles:

- `correlate_hr = TRUE` — share one `rnorm()` draw vector across all
  lognormal HR specs in an iteration. This is the legacy "single-Z"
  convention from the original Excel implementation: high-OW iterations
  also have high-OB1 and high-OB2.
- `correlate_cost = TRUE` — share one `runif()` draw across all gamma
  cost specs in an iteration ("single-U"). High-cost iterations are
  high across every state-age cell jointly.

Both default to `TRUE` to match the published runs. Set them to `FALSE`
for independent draws per spec — useful if you're investigating
how much of the credible-interval width comes from the correlation
assumption itself.

```{r psa-correlate, eval = FALSE}
spec <- moon_params_norway("female", uncertainty = TRUE)

psa_corr <- moon_psa(spec, n_iter = 100, seed = 1, correlate_hr = TRUE)
psa_indp <- moon_psa(spec, n_iter = 100, seed = 1, correlate_hr = FALSE)
```


# 13. What the validator catches (and what it doesn't)

A short bestiary of the most common errors and the messages they
produce.

| What you did | Validator says |
|---|---|
| `params$transition_probs <- params$transition_probs[1:50, ]` | `` `transition_probs$age` must cover exactly 2:99. `` |
| `params$init_prev <- c(NW = 0.5, OW = 0.5, OB1 = 0.0, OB2 = 0.0)` (typo, sums to 1.0) | (passes — sum is OK) |
| `params$init_prev <- c(NW = 0.5, OW = 0.4, OB1 = 0, OB2 = 0)` | `` `init_prev` must sum to 1 within 1e-8 (got 0.9). `` |
| `params$qx <- params$qx[1:30]` | `` `qx` is missing age(s): 32-99 (68 values). `` |
| `params$mortality_hr$OW[1] <- -0.5` | `` `mortality_hr$OW` must be positive and finite. `` |
| `params$cost_df$cost[1] <- -100` | `` `cost_df$cost` contains negative or non-finite values. `` |

What the validator **cannot** catch:

- Whether `init_prev` matches the cohort age at `start_age`.
- Whether your costs are in the currency you've labelled them.
- Whether your life table is for the population you think you're
  modelling.
- Whether transition probabilities are calibrated to the same cohort
  the costs and HRs come from.

These are domain-correctness questions; they're on you. A useful
discipline is to keep a short comment block alongside any custom
`params` construction documenting the source of each slot.


# 14. Where to go next

- **`?moon_params_norway`, `?moon_check_params`** — argument-by-argument
  reference.
- **`?moon_param_lognormal`, `?moon_param_gamma`, `?moon_param_mvnorm`,
  `?moon_param_dirichlet`, `?moon_param_fixed`** — spec constructors.
- **Getting started with MOON** — the entry-point vignette covering the
  bundled-defaults workflow.
- **Bjørnelv et al. 2021** —
  [doi:10.1177/0272989X20971589](https://doi.org/10.1177/0272989X20971589).
  Supplementary appendices 3 and 4 hold the source values for
  `mortality_hr` and `cost_df`.
